---
title: "REclaim vs external repeat tools: benchmark on simulated data"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  ext_dir: "."
  reclaim_eval_dir: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(scales)
})

ext_dir <- params$ext_dir
reclaim_eval_dir <- params$reclaim_eval_dir
```

## What this report shows

Side-by-side accuracy of REclaim's quantifiers (STARsolo unique, kallisto, alevin, bowtie2) and external tools (TEtranscripts, scTE, optionally SQuIRE) on the same simulated dataset, scored against the same per-locus ground truth at the same granularities. Every number is comparable: same fastqs, same BAMs, same locus_map, same evaluate.py.

### Native granularity coverage

REclaim is the only quantifier here that natively reports at every granularity (locus, gene_id, family_id, class_id) from a single BAM pass through the GTF feature hierarchy. The external tools each emit at one native level and rely on the same locus_map rollup that REclaim provides; rows scored at a non-native granularity are aggregated counts, not independent measurements.

| Quantifier | Native level | locus | gene_id | family_id | class_id |
|---|---|---|---|---|---|
| REclaim (STARsolo, kallisto, alevin, bowtie2) | locus / gene_id | native | native | native | native |
| TEtranscripts | gene_id (subfamily) | not produced | native | rolled up | rolled up |
| scTE | family_id | not produced | not possible | native | rolled up |
| SQuIRE | locus | native | rolled up | rolled up | rolled up |

In the panels below, REclaim bars at family_id and class_id reflect independent counts on those feature_sets, not a post-hoc rollup of gene_id counts.

```{r load-data}
ext_summary_path <- file.path(ext_dir, "summary_global_metrics.tsv")
reclaim_summary_path <- file.path(reclaim_eval_dir, "summary_global_metrics.tsv")
stopifnot(file.exists(ext_summary_path))
stopifnot(file.exists(reclaim_summary_path))

ext_df <- read_tsv(ext_summary_path, show_col_types = FALSE)
reclaim_df <- read_tsv(reclaim_summary_path, show_col_types = FALSE)

ext_df$source <- "external"
reclaim_df$source <- "reclaim"
shared_cols <- intersect(colnames(ext_df), colnames(reclaim_df))
combined <- bind_rows(
  reclaim_df[, shared_cols],
  ext_df[, shared_cols])

stopifnot(nrow(combined) > 0)
combined <- combined %>%
  mutate(across(any_of(c("pearson_r", "spearman_r", "log1p_rmse",
                         "precision", "recall", "f1", "jaccard", "specificity",
                         "wall_time_s", "max_rss_mb", "mutation_rate")),
                ~ suppressWarnings(as.numeric(.))))
```

```{r show-rows}
combined %>%
  arrange(source, aligner, granularity) %>%
  knitr::kable()
```

## Pearson r against ground truth

```{r pearson, fig.width=10, fig.height=4}
p_df <- combined %>% filter(!is.na(pearson_r))
ggplot(p_df, aes(x = aligner, y = pearson_r, fill = source)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_wrap(~ granularity, nrow = 1) +
  geom_text(aes(label = sprintf("%.3f", pearson_r)),
            position = position_dodge2(width = 0.9, preserve = "single"),
            vjust = -0.3, size = 3) +
  scale_y_continuous(limits = c(NA, 1.05)) +
  labs(x = NULL, y = "Pearson r vs ground truth",
       fill = NULL,
       title = "Per-quantifier Pearson r against ground truth, by granularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

How to read this. Bars at or near 1 are near-perfect quantifiers. The colour split (REclaim vs external) is the headline. At granularities marked "rolled up" in the table above, the external bar reflects a post-hoc aggregation through the same locus_map REclaim already used, so a REclaim win there is a win on the native quantification side, not on rollup logic.

### Headline: best quantifier per granularity

```{r headline}
headline <- combined %>%
  filter(!is.na(pearson_r)) %>%
  group_by(granularity) %>%
  arrange(desc(pearson_r)) %>%
  slice_head(n = 3) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(granularity, rank, aligner, source, pearson_r, recall, specificity)

knitr::kable(headline, digits = 3,
             caption = "Top 3 quantifiers by Pearson r at each granularity.")
```

## Sensitivity (recall)

```{r recall, fig.width=10, fig.height=4}
ggplot(combined %>% filter(!is.na(recall)),
       aes(x = aligner, y = recall, fill = source)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_wrap(~ granularity, nrow = 1) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "recall (TP / (TP+FN))",
       fill = NULL,
       title = "Per-quantifier recall, by granularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Specificity

```{r specificity, fig.width=10, fig.height=4}
ggplot(combined %>% filter(!is.na(specificity)),
       aes(x = aligner, y = specificity, fill = source)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_wrap(~ granularity, nrow = 1) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = "specificity (TN / (TN+FP))",
       fill = NULL,
       title = "Per-quantifier specificity, by granularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Precision and F1

```{r precision-f1, fig.width=10, fig.height=8}
pf_long <- combined %>%
  select(aligner, source, granularity, precision, f1) %>%
  pivot_longer(c(precision, f1), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

ggplot(pf_long, aes(x = aligner, y = value, fill = source)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(metric ~ granularity, scales = "free_y") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Per-quantifier precision and F1, by granularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Per-class breakdown

```{r per-class, fig.width=11, fig.height=6}
collect_per_class <- function(dir_path, label) {
  files <- list.files(dir_path,
                      pattern = "_per_family_metrics\\.tsv$",
                      recursive = TRUE, full.names = TRUE)
  if (!length(files)) return(NULL)
  rows <- lapply(files, function(p) {
    tryCatch(read_tsv(p, show_col_types = FALSE), error = function(e) NULL)
  })
  rows <- rows[!sapply(rows, is.null)]
  if (!length(rows)) return(NULL)
  bind_rows(rows) %>% mutate(source = label)
}

per_class <- bind_rows(
  collect_per_class(reclaim_eval_dir, "reclaim"),
  collect_per_class(ext_dir, "external"))

if (!is.null(per_class) && nrow(per_class) > 0) {
  per_class$pearson_r <- suppressWarnings(as.numeric(per_class$pearson_r))
  ggplot(per_class %>% filter(!is.na(pearson_r), !is.na(class_id)),
         aes(x = class_id, y = pearson_r, fill = source)) +
    geom_col(position = position_dodge2(preserve = "single")) +
    facet_grid(granularity ~ aligner, scales = "free_x") +
    scale_y_continuous(limits = c(NA, 1.05)) +
    labs(x = NULL, y = "Pearson r per class",
         fill = NULL,
         title = "Per-class Pearson r, faceted by quantifier and granularity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("No per-family metrics files found; skipping per-class panel.\n")
}
```

How to read this. Each panel column is one quantifier. Each subpanel row is a granularity. Within a panel, the bars are RepeatMasker classes. A quantifier that maintains high Pearson r across all classes is robust; a quantifier that collapses on a specific class points to that class being mishandled (often LTR for tools that ignore long internal sequences, or simple repeats for tools that filter low-complexity).

## Compute cost

```{r runtime, fig.width=10, fig.height=4}
cost_df <- combined %>%
  filter(!is.na(wall_time_s)) %>%
  distinct(aligner, source, wall_time_s, max_rss_mb)

if (nrow(cost_df) > 0) {
  ggplot(cost_df,
         aes(x = aligner, y = wall_time_s, fill = source)) +
    geom_col(position = position_dodge2(preserve = "single")) +
    labs(x = NULL, y = "wall-clock seconds",
         fill = NULL,
         title = "Per-tool runtime on the simulated dataset") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("No wall_time_s rows; runtime panel skipped.\n")
}
```

```{r rss, fig.width=10, fig.height=4}
if (nrow(cost_df) > 0 && any(!is.na(cost_df$max_rss_mb))) {
  ggplot(cost_df,
         aes(x = aligner, y = max_rss_mb, fill = source)) +
    geom_col(position = position_dodge2(preserve = "single")) +
    labs(x = NULL, y = "peak RSS (MB)",
         fill = NULL,
         title = "Per-tool peak memory on the simulated dataset") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  cat("No max_rss_mb rows; memory panel skipped.\n")
}
```

## Tools skipped or stubbed

```{r skipped, results='asis'}
cat("- TEtranscripts: included; bulk only; treats each SmartSeq2 cell as one sample.\n")
cat("- scTE: included; sc only; consumes the multi-cell Chromium STARsolo BAM.\n")
cat("- SQuIRE: optional; pinned to Python 3.6 and STAR 2.5.3a; may fail to install on newer machines. When unavailable, the harmoniser asserts on missing overlap and SQuIRE is dropped from this report. See docs/external_tool_benchmark.md for the rationale and the install caveats.\n")
cat("- REdiscoverTE: not included. It bundles a kallisto-based custom transcriptome that overlaps with REclaim's existing kallisto path; including it would require a parallel pseudoalignment branch outside the BAM-based comparison this report performs. Documented in docs/external_tool_benchmark.md.\n")
```

## Session info

```{r session-info}
sessionInfo()
```
