---
title: "repeat element quantification - noise sweep report"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  eval_dirs: "."
  csv_outdir: "."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(scales)
library(patchwork)
source(file.path("scripts", "ggtheme.R"))


eval_dirs <- strsplit(params$eval_dirs, ",")[[1]]
eval_dirs <- trimws(eval_dirs)
```

```{r load-data}
load_metrics_from_dirs <- function(dirs, filename) {
  purrr::map_dfr(dirs, function(d) {
    p <- file.path(d, filename)
    if (!file.exists(p)) return(NULL)
    read_tsv(p, show_col_types = FALSE)
  })
}

load_pattern_from_dirs <- function(dirs, pattern) {
  purrr::map_dfr(dirs, function(d) {
    files <- list.files(d, pattern = pattern, full.names = TRUE)
    if (length(files) == 0) return(NULL)
    purrr::map_dfr(files, ~ read_tsv(.x, show_col_types = FALSE))
  })
}

global <- load_metrics_from_dirs(eval_dirs, "summary_global_metrics.tsv")

num_cols <- c("pearson_r", "spearman_r", "log1p_rmse",
              "precision", "recall", "f1", "jaccard", "specificity",
              "n_cells", "n_features", "mutation_rate",
              "wall_time_s", "cpu_time_s", "max_rss_mb", "io_in_mb", "io_out_mb")
for (col in num_cols) {
  if (col %in% names(global))
    global[[col]] <- suppressWarnings(as.numeric(global[[col]]))
}

if (!"granularity" %in% names(global))   global$granularity   <- "gene_id"
if (!"feature_set" %in% names(global))   global$feature_set   <- "repeats"
if (!"multimapper_mode" %in% names(global)) global$multimapper_mode <- "unique"
if (!"mutation_rate" %in% names(global)) global$mutation_rate <- NA_real_

global <- global %>%
  mutate(
    aligner_mode  = paste0(aligner, " (", multimapper_mode, ")"),
    gran_label    = factor(granularity,
                           levels = c("locus", "gene_id", "family_id", "class_id")),
    noise_label   = paste0(mutation_rate * 100, "%")
  )

per_cell   <- tryCatch(load_pattern_from_dirs(eval_dirs, "_per_cell_metrics\\.tsv$"),
                       error = function(e) NULL)
per_family <- tryCatch(load_pattern_from_dirs(eval_dirs, "_per_family_metrics\\.tsv$"),
                       error = function(e) NULL)

for (col in c("pearson_r", "spearman_r", "mutation_rate")) {
  if (!is.null(per_cell) && col %in% names(per_cell))
    per_cell[[col]] <- suppressWarnings(as.numeric(per_cell[[col]]))
  if (!is.null(per_family) && col %in% names(per_family))
    per_family[[col]] <- suppressWarnings(as.numeric(per_family[[col]]))
}

if (!is.null(per_family) && nrow(per_family) > 0) {
  for (col in c("f1", "precision", "recall", "jaccard", "specificity")) {
    if (col %in% names(per_family))
      per_family[[col]] <- suppressWarnings(as.numeric(per_family[[col]]))
  }
  if (!"multimapper_mode" %in% names(per_family)) per_family$multimapper_mode <- "unique"
  per_family <- per_family %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"))
}
```

---

This report combines evaluation results across multiple noise levels (simulated nucleotide
substitution rates). Each `eval_dirs` entry corresponds to one pipeline run with a distinct
`mutation_rate` in the simulation config. The `mutation_rate` column in the metrics TSVs
identifies each run.

Noise levels found in the data: `r sort(unique(global$mutation_rate))`

---

## metric degradation by noise level

How each aligner's accuracy changes as the per-base mutation rate increases.
x-axis = mutation rate, y-axis = metric value, colour = aligner (+ multimapper mode).

```{r degradation-main, fig.height=15, fig.width=12}
acc_metrics <- c("pearson_r", "spearman_r", "f1", "precision", "recall",
                 "jaccard", "specificity")

deg <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, all_of(acc_metrics)) %>%
  pivot_longer(all_of(acc_metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(metric = factor(metric, levels = acc_metrics))


ggplot(deg, aes(x = mutation_rate, y = value,
                colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_line(linewidth = 0.6) +
  facet_grid(metric ~ feature_set + gran_label, scales = "free_y") +
  scale_colour_aligner_mode(name = "aligner (mode)") +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(x = "mutation rate", y = NULL,
       title = "metric degradation with increasing noise",
       subtitle = "rows = metric; columns = feature_set x granularity; x = mutation rate") +
  theme_ng() +
  theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
        strip.text   = element_text(size = 7),
        legend.position = "bottom",
        legend.text  = element_text(size = 8))
```

## pearson r and F1 - noise overview

Summary view focusing on the two most interpretable metrics.

```{r overview-two-metrics, fig.height=6, fig.width=14}
ov <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, pearson_r, f1) %>%
  pivot_longer(c(pearson_r, f1), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))

eps <- 1e-3
logit_clip <- scales::trans_new(
  "logit_clip",
  transform = function(x) { x <- pmin(pmax(x, eps), 1 - eps); log(x / (1 - x)) },
  inverse   = function(y) 1 / (1 + exp(-y)),
  breaks    = function(lim) c(0.5, 0.8, 0.9, 0.95, 0.99, 0.999),
  format    = function(b) format(b, drop0trailing = TRUE))

ggplot(ov, aes(x = mutation_rate, y = value,
               colour = aligner_mode,
               group = aligner_mode)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  geom_point(size = 2.2, alpha = 0.95) +
  facet_grid(metric ~ feature_set + gran_label, scales = "free_y") +
  scale_colour_aligner_mode() +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_y_continuous(trans = logit_clip) +
  labs(x = "mutation rate", y = NULL,
       title = "Pearson r and F1 across noise levels",
       subtitle = "colour = aligner_mode; columns = feature_set x granularity; y on logit scale") +
  theme_ng() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

## global accuracy metrics per noise level

Full metric grid from the standard report, faceted by mutation rate.

```{r global-faceted, fig.height=18, fig.width=14}
full_metrics <- c("pearson_r", "spearman_r", "f1", "precision", "recall",
                  "jaccard", "specificity", "log1p_rmse")

full_long <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner_mode, feature_set, gran_label, mutation_rate, noise_label,
         all_of(full_metrics)) %>%
  pivot_longer(all_of(full_metrics), names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(metric = factor(metric, levels = full_metrics))


ggplot(full_long, aes(x = gran_label, y = value,
                      colour = aligner_mode, group = aligner_mode)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(linetype = "dashed", linewidth = 0.5) +
  facet_grid(metric ~ noise_label + feature_set, scales = "free_y") +
  scale_colour_aligner_mode(name = "aligner (mode)") +
  labs(x = "granularity", y = NULL,
       title = "global accuracy metrics",
       subtitle = "rows = metric; columns = mutation rate x feature_set; x = granularity") +
  theme_ng() +
  theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
        strip.text   = element_text(size = 6),
        legend.position = "bottom",
        legend.text  = element_text(size = 7))
```

## per-cell correlations by noise level

```{r per-cell-noise, fig.height=10, fig.width=14}
if (!is.null(per_cell) && nrow(per_cell) > 0 && "mutation_rate" %in% names(per_cell)) {
  pc <- per_cell %>%
    mutate(across(c(pearson_r, spearman_r), ~ suppressWarnings(as.numeric(.)))) %>%
    filter(!is.na(aligner), !is.na(mutation_rate))

  if (!"multimapper_mode" %in% names(pc)) pc$multimapper_mode <- "unique"
  if (!"feature_set" %in% names(pc))      pc$feature_set      <- "repeats"

  pc <- pc %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"),
           gran_label   = factor(granularity,
                                 levels = c("locus", "gene_id", "family_id", "class_id")),
           noise_label  = paste0(mutation_rate * 100, "%"))

  pc_long <- pc %>%
    pivot_longer(c(pearson_r, spearman_r), names_to = "metric", values_to = "r") %>%
    filter(!is.na(r))


  ggplot(pc_long, aes(x = aligner_mode, y = r, fill = aligner_mode)) +
    geom_violin(alpha = 0.7, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(width = 0.1, size = 0.5, alpha = 0.4) +
    facet_grid(metric + gran_label ~ noise_label + feature_set) +
    scale_fill_aligner_mode(guide = "none") +
    labs(x = NULL, y = "correlation",
         title = "per-cell correlations by noise level",
         subtitle = "rows = metric x granularity; columns = mutation rate x feature_set") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 6),
          legend.position = "none",
          strip.text   = element_text(size = 6))
} else {
  cat("No per-cell metrics with mutation_rate column found.\n")
}
```

## per-repeat-class F1 by noise level

```{r per-family-noise, fig.height=14, fig.width=14}
if (!is.null(per_family) && nrow(per_family) > 0 &&
    "class_id" %in% names(per_family) && "mutation_rate" %in% names(per_family)) {

  pf <- per_family %>%
    filter(!is.na(class_id), class_id != "unknown", !is.na(mutation_rate)) %>%
    mutate(gran_label  = factor(granularity,
                                levels = c("locus", "gene_id", "family_id", "class_id")),
           noise_label = paste0(mutation_rate * 100, "%"))


  p_f1 <- ggplot(pf, aes(x = class_id, y = f1,
                          colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.4, linetype = "dashed") +
    geom_point(size = 1.5, alpha = 0.85) +
    facet_grid(gran_label ~ noise_label + feature_set) +
    scale_y_sqrt() +
    scale_colour_aligner_mode(name = "Aligner (mode)") +
    labs(x = "repeat class", y = "F1",
         title = "F1 per repeat class by noise level",
         subtitle = "rows = granularity; columns = mutation rate x feature_set") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "bottom",
          legend.text  = element_text(size = 7),
          strip.text   = element_text(size = 6))

  p_r <- ggplot(pf, aes(x = class_id, y = pearson_r,
                         colour = aligner_mode, group = aligner_mode)) +
    geom_line(linewidth = 0.4, linetype = "dashed") +
    geom_point(size = 1.5, alpha = 0.85) +
    facet_grid(gran_label ~ noise_label + feature_set) +
    scale_y_sqrt() +
    scale_colour_aligner_mode(name = "Aligner (mode)") +
    labs(x = "repeat class", y = "pearson r") +
    theme_ng() +
    theme(axis.text.x  = element_text(angle = 40, hjust = 1, size = 7),
          legend.position = "none",
          strip.text   = element_text(size = 6))

  p_f1 / p_r
} else {
  cat("No per-family metrics with class_id and mutation_rate found.\n")
}
```

## paper-ready CSV exports

Tidy CSV tables written to `{csv_outdir}/paper_csv/` for direct citation in manuscript text, tables, and figures. Each row carries `aligner`, `multimapper_mode`, `aligner_mode`, `feature_set`, `granularity`, `metric`, `mutation_rate`, and `value`, so the tables can be filtered directly for any figure panel.

```{r paper-csv-exports}
csv_outdir <- params$csv_outdir
csv_dir <- file.path(csv_outdir, "paper_csv")
dir.create(csv_dir, showWarnings = FALSE, recursive = TRUE)

write_csv_log <- function(df, name) {
  path <- file.path(csv_dir, paste0(name, ".csv"))
  write.csv(df, path, row.names = FALSE)
  cat("wrote", path, " (", nrow(df), "rows)\n")
  invisible(path)
}

higher_is_better <- c("pearson_r", "spearman_r", "f1", "precision",
                      "recall", "jaccard", "specificity")
lower_is_better <- c("log1p_rmse")
resource_cols <- intersect(
  c("wall_time_s", "cpu_time_s", "max_rss_mb", "io_in_mb", "io_out_mb"),
  names(global))

noise_long <- global %>%
  filter(!is.na(mutation_rate)) %>%
  select(aligner, multimapper_mode, aligner_mode, feature_set, granularity,
         mutation_rate,
         any_of(c(higher_is_better, lower_is_better, resource_cols,
                  "n_cells", "n_features"))) %>%
  pivot_longer(cols = -c(aligner, multimapper_mode, aligner_mode,
                         feature_set, granularity, mutation_rate),
               names_to = "metric", values_to = "value") %>%
  filter(!is.na(value))
write_csv_log(noise_long, "noise_metrics_long")

noise_wide <- noise_long %>%
  mutate(mutation_rate_pct = paste0("mut_", round(mutation_rate * 100, 1), "pct")) %>%
  select(-mutation_rate) %>%
  pivot_wider(names_from = mutation_rate_pct, values_from = value)
write_csv_log(noise_wide, "noise_metrics_wide")

fit_degradation <- function(d) {
  if (nrow(d) < 2) return(data.frame(slope = NA_real_, intercept = NA_real_,
                                     r_squared = NA_real_, n_points = nrow(d)))
  fit <- lm(value ~ mutation_rate, data = d)
  s <- summary(fit)
  data.frame(slope = unname(coef(fit)[2]),
             intercept = unname(coef(fit)[1]),
             r_squared = s$r.squared,
             n_points = nrow(d))
}
noise_degradation <- noise_long %>%
  filter(metric %in% c(higher_is_better, lower_is_better)) %>%
  group_by(aligner, multimapper_mode, aligner_mode,
           feature_set, granularity, metric) %>%
  do(fit_degradation(.)) %>%
  ungroup() %>%
  arrange(feature_set, granularity, metric, slope)
write_csv_log(noise_degradation, "noise_degradation_slopes")

robustness_ranking <- noise_degradation %>%
  filter(metric %in% higher_is_better, !is.na(slope)) %>%
  group_by(feature_set, granularity, metric) %>%
  arrange(slope, .by_group = TRUE) %>%
  mutate(rank_robustness = row_number()) %>%
  ungroup()
write_csv_log(robustness_ranking, "noise_robustness_ranking")

if (!is.null(per_cell) && nrow(per_cell) > 0 &&
    "mutation_rate" %in% names(per_cell)) {
  pc <- per_cell %>%
    mutate(across(c(pearson_r, spearman_r),
                  ~ suppressWarnings(as.numeric(.)))) %>%
    filter(!is.na(aligner), !is.na(mutation_rate))
  if (!"multimapper_mode" %in% names(pc)) pc$multimapper_mode <- "unique"
  if (!"feature_set" %in% names(pc)) pc$feature_set <- "repeats"
  pc <- pc %>%
    mutate(aligner_mode = paste0(aligner, " (", multimapper_mode, ")"))
  per_cell_noise_summary <- pc %>%
    pivot_longer(c(pearson_r, spearman_r),
                 names_to = "metric", values_to = "r") %>%
    filter(!is.na(r)) %>%
    group_by(aligner, multimapper_mode, aligner_mode,
             feature_set, granularity, mutation_rate, metric) %>%
    summarise(n_cells = dplyr::n(),
              mean_r = mean(r),
              median_r = median(r),
              sd_r = sd(r),
              q25_r = quantile(r, 0.25, na.rm = TRUE),
              q75_r = quantile(r, 0.75, na.rm = TRUE),
              .groups = "drop")
  write_csv_log(per_cell_noise_summary, "per_cell_noise_summary")
}

if (!is.null(per_family) && nrow(per_family) > 0 &&
    "class_id" %in% names(per_family) &&
    "mutation_rate" %in% names(per_family)) {
  pf_keep <- intersect(
    c("pearson_r", "spearman_r", "f1", "precision",
      "recall", "jaccard", "specificity"),
    names(per_family))
  per_class_noise <- per_family %>%
    filter(!is.na(class_id), class_id != "unknown", !is.na(mutation_rate)) %>%
    select(aligner, multimapper_mode, aligner_mode, feature_set, granularity,
           mutation_rate, class_id, all_of(pf_keep))
  write_csv_log(per_class_noise, "per_class_noise_metrics")
}
```

---
*Generated by [noise_sweep_report.Rmd](../workflow/scripts/noise_sweep_report.Rmd)*
