---
title: "Replogle 2022 K562 essential Perturb-seq alignment QC"
author: "Izaskun Mallona"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
params:
  starsolo_base: "."
  cells_to_perturbation: "."
  indices_base: "."
  report_outdir: "."
  feature_sets: "genes genic_repeats intergenic_repeats"
  starsolo_modes: "unique multi"
  starsolo_log_summary: ""
  per_cell_mismatch: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(Matrix)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(ggplot2)
})

starsolo_base <- params$starsolo_base
cells_path <- params$cells_to_perturbation
feature_sets <- strsplit(params$feature_sets, " ")[[1]]
starsolo_modes <- strsplit(params$starsolo_modes, " ")[[1]]

stopifnot(dir.exists(starsolo_base), file.exists(cells_path),
          length(feature_sets) >= 1, length(starsolo_modes) >= 1)
```

Short alignment QC for the Replogle 2022 K562 essential Perturb-seq run. Counts cells, detected genes, and detected repeats per library and per (mode, featureset). The downstream per-perturbation pseudobulk RUVg analysis lives in `paper/scripts/ruv_repogle_2022_perturbseq_report.Rmd`.

## Cells and perturbations in the analysis subset

```{r cells-summary}
cells <- read_tsv(cells_path, show_col_types = FALSE)
n_cells <- nrow(cells)
n_perturbations <- length(unique(cells$gene_transcript))
n_target_genes <- length(unique(cells$gene))
n_gemgroups <- length(unique(cells$gem_group))
cat("Cells in cells_to_perturbation.tsv:", n_cells, "\n")
cat("Distinct gene_transcript ids:    ", n_perturbations, "\n")
cat("Distinct target genes:           ", n_target_genes, "\n")
cat("Distinct gemgroups:              ", n_gemgroups, "\n")
```

## Per-library STARsolo summary

For each library and each (multimapper_mode, feature_set), counts cells with at least one UMI, total UMIs, and features with at least one UMI.

```{r per-lib-summary}
read_starsolo <- function(library_id, mode, fs) {
  mtx_dir <- file.path(starsolo_base, library_id,
                       paste0(mode, "_", fs), "Solo.out", "Gene", "raw")
  if (!dir.exists(mtx_dir)) return(NULL)
  m <- readMM(file.path(mtx_dir, "matrix.mtx"))
  barcodes <- readLines(file.path(mtx_dir, "barcodes.tsv"))
  list(matrix = m, barcodes = barcodes)
}

libraries <- sort(unique(paste0("KD6_", cells$gem_group, "_essential")))

per_lib_rows <- list()
for (lib in libraries) {
  for (mode in starsolo_modes) {
    for (fs in feature_sets) {
      r <- read_starsolo(lib, mode, fs)
      if (is.null(r)) next
      cells_with_umi <- sum(colSums(r$matrix) > 0)
      features_detected <- sum(rowSums(r$matrix) > 0)
      total_umi <- sum(r$matrix)
      per_lib_rows[[length(per_lib_rows) + 1]] <- data.frame(
        library = lib, mode = mode, feature_set = fs,
        n_barcodes = length(r$barcodes),
        n_cells_with_umi = cells_with_umi,
        n_features_detected = features_detected,
        total_umi = total_umi,
        stringsAsFactors = FALSE)
    }
  }
}
per_lib <- bind_rows(per_lib_rows)
kable(per_lib)
```

## Aggregated totals

Cells, genes, and repeat features detected across all libraries (union) per (mode, featureset).

```{r totals}
totals <- per_lib %>%
  group_by(mode, feature_set) %>%
  summarise(
    libraries = n(),
    total_barcodes = sum(n_barcodes),
    total_cells_with_umi = sum(n_cells_with_umi),
    total_umi = sum(total_umi),
    .groups = "drop")
kable(totals)
```

## Cells matched to the perturbation table

How many of the STARsolo barcodes per library appear in `cells_to_perturbation.tsv`. Mismatches indicate cells filtered out by the Phase 1 selection or barcode-format mismatches.

```{r cells-matched}
match_rows <- list()
for (lib in libraries) {
  gg <- as.integer(sub("^KD6_(\\d+)_essential$", "\\1", lib))
  expected_bc <- sub("-[0-9]+$", "", cells$cell_barcode[cells$gem_group == gg])
  for (mode in starsolo_modes) {
    fs <- feature_sets[1]
    r <- read_starsolo(lib, mode, fs)
    if (is.null(r)) next
    matched <- sum(expected_bc %in% r$barcodes)
    match_rows[[length(match_rows) + 1]] <- data.frame(
      library = lib, mode = mode,
      n_expected = length(expected_bc),
      n_matched = matched,
      pct_matched = round(100 * matched / max(1, length(expected_bc)), 1),
      stringsAsFactors = FALSE)
  }
}
matches <- bind_rows(match_rows)
kable(matches)
```

## Repeat feature totals

Distinct repeat features detected (union across libraries) per (mode, repeat featureset). Compares to gene featureset for context.

```{r repeat-totals}
union_rows <- list()
for (mode in starsolo_modes) {
  for (fs in feature_sets) {
    detected <- character(0)
    for (lib in libraries) {
      r <- read_starsolo(lib, mode, fs)
      if (is.null(r)) next
      mtx_dir <- file.path(starsolo_base, lib,
                            paste0(mode, "_", fs), "Solo.out", "Gene", "raw")
      features <- readLines(file.path(mtx_dir, "features.tsv"))
      features <- sapply(strsplit(features, "\t"), `[`, 1)
      detected <- union(detected,
                         features[rowSums(r$matrix) > 0])
    }
    union_rows[[length(union_rows) + 1]] <- data.frame(
      mode = mode, feature_set = fs,
      n_features_union = length(detected),
      stringsAsFactors = FALSE)
  }
}
unions <- bind_rows(union_rows)
kable(unions)
```

## STAR log variability

Mismatch and indel rates per library, parsed from `Log.final.out`. Differences between libraries can point to sequencing or library quality differences.

```{r starsolo-log-variability, fig.width = 10, fig.height = 6}
log_path <- params$starsolo_log_summary
if (nzchar(log_path) && file.exists(log_path)) {
  star_log <- read_tsv(log_path, show_col_types = FALSE)
  saveRDS(star_log, file.path(params$report_outdir, "starsolo_log_summary.rds"))

  numeric_cols <- c("mismatch_rate_pct", "deletion_rate_pct",
                    "insertion_rate_pct", "uniquely_mapped_pct",
                    "multi_mapped_pct", "too_many_loci_pct",
                    "unmapped_too_short_pct", "unmapped_other_pct")
  star_log <- star_log %>%
    mutate(across(any_of(numeric_cols),
                  ~ suppressWarnings(as.numeric(.))))

  rate_long <- star_log %>%
    select(sample_id, multimapper_mode, feature_set,
           mismatch_rate_pct, deletion_rate_pct, insertion_rate_pct) %>%
    pivot_longer(c(mismatch_rate_pct, deletion_rate_pct, insertion_rate_pct),
                 names_to = "metric", values_to = "pct") %>%
    filter(!is.na(pct))

  print(ggplot(rate_long, aes(x = sample_id, y = pct, colour = feature_set,
                              shape = multimapper_mode)) +
    geom_point(size = 2.2, alpha = 0.85) +
    facet_wrap(~ metric, scales = "free_y", ncol = 1) +
    labs(x = NULL, y = "rate (%)",
         title = "Per-library STAR mismatch and indel rates",
         subtitle = "one point per (library, mode, feature_set)") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          legend.position = "bottom"))

  map_long <- star_log %>%
    select(sample_id, multimapper_mode, feature_set,
           uniquely_mapped_pct, multi_mapped_pct,
           too_many_loci_pct, unmapped_too_short_pct, unmapped_other_pct) %>%
    pivot_longer(-c(sample_id, multimapper_mode, feature_set),
                 names_to = "fate", values_to = "pct") %>%
    filter(!is.na(pct))

  print(ggplot(map_long, aes(x = sample_id, y = pct, fill = fate)) +
    geom_col(position = "stack") +
    facet_grid(multimapper_mode ~ feature_set) +
    labs(x = NULL, y = "% reads",
         title = "Per-library STAR read-fate breakdown") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
          legend.position = "bottom"))

  star_log %>%
    group_by(multimapper_mode, feature_set) %>%
    summarise(across(any_of(numeric_cols),
                     list(mean = ~ mean(., na.rm = TRUE),
                          sd   = ~ sd(., na.rm = TRUE)),
                     .names = "{.col}_{.fn}"),
              .groups = "drop") %>%
    kable(digits = 3,
          caption = "STAR log metric summary across libraries (mean +/- sd)")
} else {
  cat("starsolo_log_summary TSV not provided or not found:", log_path, "\n")
}
```

## Per-cell mismatch rate by perturbation

Per-cell `NM/aligned_bases` from each library's BAM, joined to the perturbation table by `(gem_group, cell_barcode)`. Each guide pair (`sgID_AB`) targeting the same gene gives one summary point, so genes with multiple guide pairs show the within-gene spread.

```{r per-cell-mismatch, fig.width = 11, fig.height = 7}
mm_path <- params$per_cell_mismatch
if (nzchar(mm_path) && file.exists(mm_path)) {
  mm <- read_tsv(mm_path, show_col_types = FALSE)
  cells_full <- read_tsv(cells_path, show_col_types = FALSE) %>%
    mutate(cb_short = sub("-[0-9]+$", "", cell_barcode))

  joined <- mm %>%
    inner_join(cells_full,
               by = c("gem_group" = "gem_group",
                      "cell_barcode" = "cb_short"))
  saveRDS(joined, file.path(params$report_outdir, "per_cell_mismatch.rds"))

  ntc_baseline <- joined %>%
    filter(gene == "non-targeting") %>%
    summarise(median_rate = median(mismatch_rate, na.rm = TRUE)) %>%
    pull(median_rate)

  per_guide <- joined %>%
    group_by(gene, sgID_AB) %>%
    summarise(n_cells = dplyr::n(),
              median_rate = median(mismatch_rate, na.rm = TRUE),
              q25 = quantile(mismatch_rate, 0.25, na.rm = TRUE),
              q75 = quantile(mismatch_rate, 0.75, na.rm = TRUE),
              .groups = "drop") %>%
    filter(n_cells >= 5)

  per_gene <- per_guide %>%
    group_by(gene) %>%
    summarise(n_guides = dplyr::n(),
              gene_median = median(median_rate),
              gene_min = min(median_rate),
              gene_max = max(median_rate),
              .groups = "drop")

  multi_guide_genes <- per_gene %>% filter(n_guides >= 2) %>% pull(gene)

  print(ggplot(per_guide %>% filter(gene %in% multi_guide_genes |
                                    gene == "non-targeting"),
               aes(x = reorder(gene, median_rate, FUN = median),
                   y = median_rate)) +
    geom_hline(yintercept = ntc_baseline,
               linetype = "dashed", colour = "grey40") +
    geom_point(aes(size = n_cells, colour = gene == "non-targeting"),
               alpha = 0.75) +
    scale_colour_manual(values = c(`TRUE` = "grey40", `FALSE` = "steelblue"),
                        guide = "none") +
    coord_flip() +
    labs(x = NULL, y = "median per-cell mismatch rate (NM / aligned bases)",
         title = "Mismatch rate by perturbation",
         subtitle = sprintf("one point per guide pair; dashed = non-targeting median (%.4f)",
                            ntc_baseline)) +
    theme_minimal(base_size = 10))

  per_gene %>%
    arrange(desc(gene_median)) %>%
    head(25) %>%
    kable(digits = 5,
          caption = "Top 25 genes by median per-guide mismatch rate")
} else {
  cat("per_cell_mismatch TSV not provided or not found:", mm_path, "\n")
}
```
